### 2012-10-29 paper_draft.tex

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Genome-wide association studies are widely used to correlate phenotypic traits with genetic variants. These studies usually compare the genetic variation between two groups to single out certain Single Nucleotide Polymorphisms (SNPs) that are linked to a phenotypic variation in one of the groups. However, it is necessary to have a large enough sample size to find statistically significant correlations. Direct-To-Consumer (DTC) genetic testing can supply additional data: DTC-companies offer the analysis of a large amount of SNPs for an individual at low cost without the need to consult a physician or geneticist. Over 100,000 people have already been genotyped through Direct-To-Consumer genetic testing companies. However, this data is not public for a variety of reasons and thus cannot be used in research. It seems reasonable to create a central open data-repository for such data, but it was previously unknown if and how people would submit their data to such a repository. Here we present a survey which evaluates whether people are willing to publicly share their genetic information. In the light of those results we present the web-platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publicly publish at no cost their genetic data along with phenotypic information. Through this crowdsourced effort of collecting genetic and phenotypic information openSNP has become a valuable resource which can be used in a wide area of studies, including Genome-wide association studies. OpenSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr

@@

The availability of new DNA sequencing techniques has shifted the focus of biological data acquisition towards new biomedical applications.
Many diseases – for example Alzheimer's \cite{alzheimer}, Parkinson's \cite{parkinsons} or different types of cancers \cite{breastcancer,prostatecancer} – are at least partially heritable so the genome 
of patients can be used for diagnostic purposes. Using the genetic information of patients for diagnostics is made possible through the sharp decrease in costs for analysing genetic information \cite{Brown1999}.

@@

If genetic information for more than one individual is known, the analysis of
allele frequencies of Single Nucleotide Polymorphisms (SNPs) can be used to associate such SNPs with diseases and other inheritable traits. Genome-Wide Association Studies
(GWAS) make use of statistics to compare the allele frequencies in patients to the alleles in healthy controls. This
enables GWAS to find SNPs which are significantly overrepresented in patients and associates those SNPs with a trait or disease.
This method does not allow inference of causal differences but merely identifies correlations. 
The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
such studies has been rising. To date, over 1200 GWAS have been performed \cite{Johnson2009} and over 
5000 SNPs have been linked to different diseases and traits \cite{Hindorff2009}.

GWAS are not only performed inside the traditional scientific community. 
Since 2006, companies like 23andMe, deCODEme or FamilyTreeDNA have been offering Direct-To-Consumer (DTC) genetic testing. 
These companies use DNA microarrays to screen for around 0.5 to 1 million SNPs spread over the human genome. In return, customers 
receive an analysis of the results, as well as a file that includes the customer's raw individual genotypes. In 2011, 23andMe 
alone had over 100,000 customers \cite{23andMe2011}
- the company realizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about 
traits and diseases. With the consent of the customer the data is used for association studies. 23andMe has published several 
articles in which known findings are replicated together with new associations disorders like Parkinson's Disease \cite{Eriksson2010, Do2011}. 
So far, over 30,000 23andme-customers have participated in 23andme's association studies, which proves that this data-source has a lot of potential for other researchers.

@@

Here, we present the results of a survey designed to evaluate the support in the personal genetics community for a crowdsourced online platform. We also present openSNP, an online platform which enables DTC customers to share genotypic and phenoytypic information, as well as receive additional information on their genotypes. The genotypes are made available to researchers via the open Creative-Commons-Zero license.

% Results and Discussion can be combined.
\section*{Results}
\subsection*{Survey on Sharing Genetic Information}
In total 229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX participated in our survey on sharing genetic information with the public. 
The mean age of the participants is 33 (SD = 11,29). 81.7 \% reported their ethnicity as caucasian. 39.7 \% of the participants are already 
customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future. 29.7 \% do not plan on 
becoming a DTC customer. There is no significant difference in the usage of DTC companies between chromosomal sexes (Cramer's V = 0.077).

67.7 \% of all participants would share their data with their DTC-company without any constraints, 25.8 \% would do so given the company 
didn't share the data with third parties. 6.6 \% of the participants would not share their data. Participants self-identified as XX-chromosomal are slightly more likely to answer that DTC companies are allowed to use their results (Cramer's V = 0.221). Those who are customers of a DTC company or are planning on becoming one in 
the future are more likely to share their results, compared to those who do not plan on getting themselves genotyped (Somers-d = 0.331).

@@

Similarly, those people who would share data with their DTC provider under any circumstances are likely to agree more strongly with 
the following motivations for sharing than those who would not share their data with their DTC company.
Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the some motivations 
for not sharing their data when compared to those who would share their data with their DTC company under any circumstances, for an overview of the motivations of both groups, see table \ref{tab:motivations2}.

@@

These survey-results indicate that there is a definite interest in customers of DTC-companies to share their results with other scientists.

@@

We created the openSNP project (http://opensnp.org) as an open, crowdsourced online platform for DTC customers interested in sharing their raw data and for researchers interested in performing GWAS or perform other types of analysis with the data. 
Customers of DTC testing are encouraged to share their genotyping results along with their phenotypic traits to enable easy access for researchers.
Users of openSNP can create a personal profile, discuss SNPs and phenotypes on the platform using a simple commenting system, or send each other private messages.

@@

\subsection*{Sharing genotypic information}
Currently users can upload their genotyping results from the companies \textit{23andMe}, \textit{deCODEme }and \textit{FamilyTreeDNA} via a webinterface to the openSNP 
project. There is experimental support for uploading exomes in the VCF format \cite{Danecek01082011}, as \textit{23andMe} recently started exome sequencing for its customers. So far only the SNPs of the exome data sets are visualized on openSNP, but the downloads include all variation found in the exome.
The uploaded data is published under the Creative Commons Zero-license, 
which - in accordance with the Panton Principles \cite{10.1371/journal.pbio.1001195} - 
allows a complete reuse of the data without any constraints.
Between the start of openSNP on 09/27/2011 and 10/27/2012, 633 people have signed 
up with openSNP, and 270 genetic datasets were made available. The openSNP 
database lists 215,546,685 genotypes which are distributed over 2,140,643 unique SNPs.
Figures \ref{Figure1_label} and \ref{Figure2_label} depicts the increase of users and genotyping files over time.

\subsection*{Crowdsourcing phenotypes}
Users are able to create new phenotypes that are not yet 
listed by openSNP. 
The specification of these phenotypes is open and not limited 
to pre-defined categories. To reduce the amount of manual data curation, openSNP tries to harmonize 
the expression and spelling of the same phenotype or variation. We implemented an 
autocompletion-feature, which helps users to reuse already entered phenotypes.
Users are encouraged to list as many phenotypes as possible through a simple 
achievement system, rewarding users that upload their data and enter phenotypic 
information with small badges that are shown on their profile pages.

In the same timeframe as above, all users combined have 
entered a total of 4743 variations on 130 different phenotypes with those variations being 
the different values on a given trait or phenotype. The mean number of users that have entered their variations for a single phenotype 
is 36.48. The distribution of how many users have 
entered their data per phenotype, compared to the amount of unique phenotypes, can be seen in figure \ref{pheno}. The phenotype provided by the most users is "eye color", for which 207 users entered their phenotype (retrieved 10/27/2012).

@@

Users are also able to link their Fitbit-accounts to their user-accounts. Fitbit is a commercial service which lets their customers track their BMI, movement-data and sleep data. This data can be linked to openSNP to give interested researchers an automatically maintained dataset of body- and sleep-developments over time.

\subsection*{Data access}
OpenSNP offers extensive access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files 
grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypic variation in a comma-separated table. The genetic data is also 
accessible through the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}, which offers all data for specific chromosomes and specific positions on single chromosomes. 
An example of how the DAS can be used can be found on openSNP where users genotypes are visualized inside a genome browser. So far all chromosomal positions are based on the human reference genome NCBI36, as this is the standard reference used by DTC providers right now. \bastian{should this maybe include an image as well?}

The data is additionally available over a JSON-API, which allows users to directly access data in the JSON-format. The methods allow users to programmatically look for the genotypes and annotations at a given SNP as well as for phenotypes for a given user and phenotypic variation for a given phenotype.

@@

%The advent of DTC genetic testing has led to new ethical and social issues. 
Much of the critique on DTC genetic testing focusses on the practice 
of delivering medical information without consulting a physician or genetic counsellor to help patients/customers make sense of the information 
and to put the new knowledge to good use \cite{Hauskeller2011,Hogarth2008,Wasson2009}.

@@

OpenSNP openly addresses the problem of privacy implications that come with releasing genetic data twice, once during registration for openSNP and once during 
the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects 
of publishing their data. Further versions of openSNP may optionally include further consent-processes.

\subsection*{GWAS and Open Data}
Although prices of exome or even full genome sequencing are dropping rapidly, GWAS are still considerably cheaper. However, GWAS can only detect correlations of SNPs with those traits and do not allow 
inference on the cause for any correlation. Furthermore, for a statistically sound analysis GWAS need a large enough sample size. Nevertheless, GWAS are still frequently used and new associations are found \cite{10.1371.journal.pone.0031470,10.1371.journal.pone.0030309,10.1371.journal.pone.0029848}.

One way of bringing down costs for GWAS even further is to make use of already available genotyping results and datasets. 
Data produced by DTC genetic testing companies is a promising source for such data, as such companies already have high 
numbers of customers which are willing to pay for the genotyping by themselves.

@@

\subsection*{Survey on Sharing Genetic Information}
The survey was performed using \textit{Google Docs} and was distributed to possible participants through the \textit{23andMe }community forums, the \textit{DIYBiology} mailing list, 
blogs which focus on genetics and DTC genetic testing and social media websites flike \textit{Twitter}, \textit{Google+} and \textit{Facebook}.

The survey included demographics such as age, chromosomal sex and ethnicity of the participants. Furthermore, it included questions on their 
(planned) customer-ship with a DTC company. If the participants already were customers, they were also asked if they were already sharing their genetic and phenotypic data. 
All participants were asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, 
but only if they did not share my medical information with anybody else" and No".

The survey also asked some scaled questions which measured how strongly participants agreed/disagreed with different reasons for sharing or not sharing their 
information. The scale went from 1 = strongly disagree to  5 = strongly agree. Motivations queried for sharing data 
were "because I want to help scientists with their research", "because of possible personal benefits (e.g. getting treatments for a disease I have, 
possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity". Motivations queried for not sharing 
data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", 
"because of the breach of my privacy" and "because of the fear of discrimination (e.g. by the employer, the state, some insurance company)". 
Additionally, participants had the possibility of giving their own reasons for sharing or not sharing their data.

@@

OpenSNP only serves as a platform for SNPs, so methods for the delivery of nucleotide sequences as described in the DAS-standard are not implemented. Currently, 
two methods are implemented: \textit{features}, which is used to deliver SNPs located on specific chromosomes or between specific nucleotide positions, 
based on the user's query. The second method is \textit{sources}, which advertises all DAS-sources for all genotypes present in openSNP.

A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in figure \ref{Figure4_label}. The source code of openSNP is 
published under the MIT-license and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
We thank Dr. Manuel Corpas and Prof. Dr. Dave Edwards for constructive advice in grammar, spelling and structure of this study. Further thanks go to Samantha Clark and Dan Bolser for providing valuable feedback and feature ideas during the development and Thomas Down, author of \emph{BioDalliance}, and Rafael Jimenez, author of \emph{MyKaryoView}, for their support on implementing the Distributed Annotation System and the genome browser. For help with their APIs we are grateful to Mike Cariaso of \emph{SNPedia} and the PLOS \& Mendeley API teams. We would especially like to thank the users of openSNP.org for their participation, their constructive criticisms and bug-finding abilities and especially for sharing their genotyping and phenotype data.

>>>>

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Genome-wide association studies are widely used to correlate phenotypic traits with genetic variants. These studies usually compare the genetic variation between two groups to single out certain Single Nucleotide Polymorphisms (SNPs) that are linked to a phenotypic variation in one of the groups. However, it is necessary to have a large enough sample size to find statistically significant correlations. Direct-To-Consumer (DTC) genetic testing can supply additional data: DTC-companies offer the analysis of a large amount of SNPs for an individual at low cost without the need to consult a physician or geneticist. Over 100,000 people have already been genotyped through Direct-To-Consumer genetic testing companies. However, this data is not public for a variety of reasons and thus cannot be used in research. It seems reasonable to create a central open data repository for such data, but it was previously unknown if and how people would submit their data to such a repository. Here we present a survey which evaluates whether people are willing to publicly share their genetic information. In the light of those results we present the web platform openSNP, an open database which allows participants of Direct-To-Consumer genetic testing to publish at no cost their genetic data along with phenotypic information. Through this crowdsourced effort of collecting genetic and phenotypic information, openSNP has become a valuable resource, which can be used in a wide area of studies, including Genome-wide association studies. OpenSNP is hosted at http://www.opensnp.org, and the code is released under MIT-license at http://github.com/gedankenstuecke/snpr

@@

The availability of new DNA sequencing techniques has shifted the focus of biological data acquisition towards new biomedical applications.
Many diseases - for example Alzheimer's \cite{alzheimer}, Parkinson's \cite{parkinsons} or different types of cancers \cite{breastcancer,prostatecancer} - are at least partially heritable, so the genome 
of patients can be used for diagnostic purposes. Using the genetic information of patients for diagnostics is made possible through the sharp decrease in costs for analysing genetic information \cite{Brown1999}.

@@

If genetic information on more than one individual is known, the analysis of
allele frequencies of Single Nucleotide Polymorphisms (SNPs) can be used to associate such SNPs with diseases and other inheritable traits. Genome-Wide Association Studies
(GWAS) make use of statistics to compare the allele frequencies in patients to the alleles in healthy controls. This
enables GWAS to find SNPs which are significantly overrepresented in patients and associates those SNPs with a trait or disease.
This method does not allow inference of causal differences but merely identifies correlations. 
The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
such studies has been rising. To date, over 1200 GWAS have been performed \cite{Johnson2009} and over 
5000 SNPs have been linked to different diseases and traits \cite{Hindorff2009}.

GWAS are not only performed inside the traditional scientific community. 
Since 2006, companies like 23andMe, deCODEme or FamilyTreeDNA have been offering Direct-To-Consumer (DTC) genetic testing. 
These companies use DNA microarrays to screen for around 0.5 to 1 million SNPs spread over the human genome. In return, customers 
receive an analysis of the results, as well as a file that includes the customer's raw individual genotypes. In 2011, 23andMe 
alone had over 100,000 customers \cite{23andMe2011}
- the company realizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about 
traits and diseases. With the consent of the customer, the data is used for association studies. 23andMe has published several 
articles in which known findings are replicated together with new associations disorders like Parkinson's Disease \cite{Eriksson2010, Do2011}. 
So far, over 30,000 23andMe-customers have participated in 23andMe's association studies, which proves that this data source has a lot of potential for other researchers.

@@

Here, we present the results of a survey designed to evaluate the support in the personal genetics community for a crowdsourced online platform. We also present openSNP, an online platform which enables DTC customers to share genotypic and phenoytypic information, as well as receive additional information on their genotypes. The genotypes are made available to researchers via the open Creative Commons Zero license.

% Results and Discussion can be combined.
\section*{Results}
\subsection*{Survey on Sharing Genetic Information}
In total 229 people, 180 with a self-reported chromosomal sex of XY and 56 with a self-reported chromosomal sex of XX, participated in our survey on sharing genetic information with the public. 
The mean age of the participants is 33 (SD = 11,29). 81.7 \% reported their ethnicity as caucasian. 39.7 \% of the participants are already 
customers of at least one DTC genetic testing company and further 30.1 \% of them plan on becoming one in the future. 29.7 \% do not plan on 
becoming a DTC customer. There is no significant difference in the usage of DTC companies between chromosomal sexes (Cramer's V = 0.077).

67.7 \% of all participants would share their data with their DTC company without any constraints, 25.8 \% would do so given the company 
didn't share the data with third parties. 6.6 \% of the participants would not share their data. Participants self-identified as XX-chromosomal are slightly more likely to answer that DTC companies are allowed to use their results (Cramer's V = 0.221). Those who are customers of a DTC company or are planning on becoming one in 
the future are more likely to share their results, compared to those who do not plan on getting themselves genotyped (Somers-d = 0.331).

@@

Similarly, those people who would share data with their DTC provider under any circumstances are likely to agree more strongly with 
the following motivations for sharing than those who would not share their data with their DTC company.
Those participants who are not willing to share data with their DTC company are likely to agree more strongly with some motivations 
for not sharing their data when compared to those who would share their data with their DTC company under any circumstances. For an overview of the motivations of both groups, see table \ref{tab:motivations2}.

@@

These survey results indicate that there is a definite interest in customers of DTC companies to share their results with other scientists.

@@

We created the openSNP project (http://opensnp.org) as an open, crowdsourced online platform for DTC customers interested in sharing their raw data and for researchers interested in performing GWAS or other types of analysis with the data. 
Customers of DTC testing are encouraged to share their genotyping results along with their phenotypic traits to enable easy access for researchers.
Users of openSNP can create a personal profile, discuss SNPs and phenotypes on the platform using a simple commenting system, or send each other private messages.

@@

\subsection*{Sharing genotypic information}
Currently users can upload their genotyping results from the companies \textit{23andMe}, \textit{deCODEme }and \textit{FamilyTreeDNA} via a web interface to the openSNP 
project. There is experimental support for uploading exomes in the VCF format \cite{Danecek01082011}, as \textit{23andMe} recently started exome sequencing for its customers. So far only the SNPs of the exome data sets are visualized on openSNP, but the downloads include all variation found in the exome.
The uploaded data is published under the Creative Commons Zero license, 
which - in accordance with the Panton Principles \cite{10.1371/journal.pbio.1001195} - 
allows a complete re-use of the data without any constraints.
Between the start of openSNP on 09/27/2011 and 10/27/2012, 633 people have signed 
up with openSNP, and 270 genetic datasets have been made available. The openSNP 
database lists 215,546,685 genotypes which are distributed over 2,140,643 unique SNPs.
Figures \ref{Figure1_label} and \ref{Figure2_label} depict the increase in users and genotyping files over time.

\subsection*{Crowdsourcing phenotypes}
Users are able to create new phenotypes that are not yet 
listed by openSNP. 
The specification of these phenotypes is open and not limited 
to pre-defined categories. To reduce the amount of manual data curation, openSNP tries to harmonize 
the expression and spelling of the same phenotype or variation. We implemented an 
autocompletion feature, which helps users reuse already entered phenotypes.
Users are encouraged to list as many phenotypes as possible through a simple 
achievement system, rewarding users that upload their data and enter phenotypic 
information with badges that are shown on their profile pages.

In the same timeframe mentioned above, all users combined have 
entered a total of 4743 variations on 130 different phenotypes with those variations being 
the different values on a given trait or phenotype. The mean number of users that have entered their variations for a single phenotype 
is 36.48. The distribution of how many users have 
entered their data per phenotype, compared to the amount of unique phenotypes, can be seen in figure \ref{pheno}. The phenotype provided by the most users is "eye color", for which 207 users entered their phenotype (retrieved 10/27/2012).

@@

Users are also able to link their Fitbit accounts to their user-accounts. Fitbit is a commercial service which lets its customers track their BMI, movement and sleep data. This data can be linked to openSNP to give interested researchers an automatically maintained dataset of body and sleep developments over time.

\subsection*{Data access}
OpenSNP offers extensive access to the data uploaded by users. Anyone can download single genotyping files for specific users, get archives of multiple genotyping files 
grouped by phenotypic variation, or access a single download that includes all genotyping files and all phenotypic variation in a comma-separated table. The genetic data is also 
accessible through the Distributed Annotation System \cite{Dowell2001,Jenkinson2008}, which offers all data for specific chromosomes and specific positions on single chromosomes. 
An example of how the DAS can be used is implemented on openSNP, where users' genotypes are visualized inside a genome browser. So far, all chromosomal positions are based on the human reference genome NCBI36, as this is the standard reference used by DTC providers right now. \bastian{should this maybe include an image as well?}

The data is additionally available over a JSON API, which allows users to directly access data in the JSON format. The methods allow users to programmatically look for the genotypes and annotations at a given SNP as well as for phenotypes for a given user and phenotypic variation for a given phenotype.

@@

%The advent of DTC genetic testing has led to new ethical and social issues. 
Much of the criticism of DTC genetic testing focusses on the practice 
of delivering medical information without consulting a physician or genetic counsellor to help patients/customers make sense of the information 
and to put the new knowledge to good use \cite{Hauskeller2011,Hogarth2008,Wasson2009}.

@@

OpenSNP openly addresses the problem of privacy implications that come with releasing genetic data twice, once during registration for openSNP and once during 
the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects 
of publishing their data. Further versions of openSNP may optionally include further consent processes.

\subsection*{GWAS and Open Data}
Although prices of exome or even full genome sequencing are dropping rapidly, GWAS are still considerably cheaper. However, GWAS can only detect correlations of SNPs with those traits and do not allow 
inference on the cause for any correlation. Furthermore, for a statistically sound analysis, GWAS need a large enough sample size. Nevertheless, GWAS are still frequently used and new associations are found \cite{10.1371.journal.pone.0031470,10.1371.journal.pone.0030309,10.1371.journal.pone.0029848}.

One way of bringing down costs for GWAS even further is to make use of already available genotyping results and datasets. 
Data produced by DTC genetic testing companies is a promising source for such results, as those companies already have high 
numbers of customers which are willing to pay for the genotyping by themselves.

@@

\subsection*{Survey on Sharing Genetic Information}
The survey was performed using \textit{Google Docs} and was distributed to possible participants through the \textit{23andMe }community forums, the \textit{DIYBiology} mailing list, 
blogs which focus on genetics and DTC genetic testing and social media websites like \textit{Twitter}, \textit{Google+} and \textit{Facebook}.

The survey included demographics such as age, chromosomal sex and ethnicity of the participants. Furthermore, it included questions on their 
(planned) customership with a DTC company. If the participants already were customers, they were also asked if they were already sharing their genetic and phenotypic data. 
All participants were asked if they would be willing to share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, 
but only if they did not share my medical information with anybody else" and No".

The survey also asked some scaled questions, which measured how strongly participants agreed/disagreed with different reasons for sharing or not sharing their 
information. The scale went from 1 = strongly disagree to  5 = strongly agree. Motivations queried for sharing data 
were "because I want to help scientists with their research", "because of possible personal benefits (e.g. getting treatments for a disease I have, 
possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity". Motivations queried for not sharing 
data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", 
"because of the breach of my privacy" and "because of the fear of discrimination (e.g. by the employer, the state, some insurance company)". 
Additionally, participants had the possibility of giving their own reasons for sharing or not sharing their data.

@@

OpenSNP only serves as a platform for SNPs, so methods for the delivery of nucleotide sequences as described in the DAS-standard are not implemented. Currently, 
two methods are implemented: firstly \textit{features}, which is used to deliver SNPs located on specific chromosomes or between specific nucleotide positions, 
based on the user's query. The second method is \textit{sources}, which advertises all DAS sources for all genotypes present in openSNP.

A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in figure \ref{Figure4_label}. The source code of openSNP is 
published under the MIT license and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
We thank Dr. Manuel Corpas and Prof. Dr. Dave Edwards for constructive advice in grammar, spelling and structure of this study. Further thanks go to Samantha Clark and Dan Bolser for providing valuable feedback and feature ideas during the development and Thomas Down, author of \emph{BioDalliance}, and Rafael Jimenez, author of \emph{MyKaryoView}, for their support on implementing the Distributed Annotation System and the genome browser. For help with their APIs we are grateful to Mike Cariaso of \emph{SNPedia} and the PLOS \& Mendeley API teams. We would especially like to thank the users of openSNP.org for their participation, their constructive criticism and bug-finding abilities and especially for sharing their genotyping and phenotype data.
